---
title: "Mini Data-Analysis Deliverable 1 - Jessica Mukiri"
author: "Jessica Mukiri"
date: "06/10/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mini Data-Analysis Deliverable 1

This is my deliverable for the mini data analysis. My tasks are to:

1. Choose favorite dataset (10 points)
2. Explore chosen dataset (15 points)
3. Write research questions (5 points)

Before starting on this task, I first need to download and load the appropriate packages

```{r }
# Packages needed
#install any of the loaded packages if not available e.g install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")

library(datateachr)
library(tidyverse)
library(ggplot2)
library(lubridate)
```

# 1 Choosing favorite dataset

## 1.1 Selecting four data sets

The criteria for selecting the data sets:

 a. No human related data
 b. Include built environment (trees,rivers,buildings)
 c. Be in Canada, a city, with a preference for Vancouver BC, then the rest of Canada
 
Based on that criteria I was able to narrow my selection to the four data sets below: 

__1. vancouver_trees__

__2. parking_meters__

__3. apt_buildings__

**4. flow_sample**

## 1.2 Exploring the 4 data sets 

In this next step I want to have a better sense of how many columns and rows the data had and what kind of data is available. 

```{r echo=FALSE}
#dplyr function glimpse was used 

glimpse(vancouver_trees)
# this data has 146,611 data rows,and 20 columns.the biggest data set I have chosen, interesting to see the very detailed data on spps name, location, diameter.(date, numeric and string data)
glimpse(flow_sample)
#smallest of all the data sets chosen, only with 218 rows and 7 columns mostly (numeric and string)
glimpse(parking_meters)
#10,032 rows and 22 columns, numeric and string data. Information on the location, payment method collected
glimpse(apt_buildings)
#3455 rows, 37 columns, numeric and string data information on type of building, year built present 
```
## 1.3 Selecting two data sets. 
Based on my initial criteria and having a look at the data I have narrowed my choices to:

**1.vancouver_trees**

**2.parking_meters**

As a new comer to Vancouver analyzing this data, will give me a better understanding on Vancouver, where trees and parking meters are located and what access I have them, which neighborhoods. 

```{r}
#seeing the different neighborhoods that trees and parking meters are located 
loc_tree <- vancouver_trees %>% 
              group_by(neighbourhood_name)  %>% 
              tally()  
loc_tree

loc_parkingm <- parking_meters  %>% 
              group_by(geo_local_area)  %>% 
              tally()  
loc_parkingm
```

## 1.4 Selecting one data set

**1.vancouver_trees**

info on [vancouver_trees](https://opendata.vancouver.ca/explore/dataset/street-trees/information/?disjunctive.species_name&disjunctive.common_name&disjunctive.height_range_id)

Vancouver is a very green city and during my first month here I have seen a lot of trees. As an environmentalist it will be interesting to explore a number of questions such as what species of trees are planted,in which neighborhoods, how old the trees are, which years were trees planted and are there certain years more trees were planted?

# 2 Exploring vancouver_trees data set

In this process I have delve into the data and have better understanding of what is available.

## 2.1a Plotting the distribution of tree diameter

This distribution will give me a sense of how wide the trees are in Vancouver.

```{r}

# first I rename the Vancouver_trees data set to make it easier when running analysis 
van_trees <- vancouver_trees

#plotting a histogram
hist_vant <- ggplot(van_trees, aes(x=diameter)) +
  geom_histogram(bins = 30)

hist_vant

# the histogram has many extremes, I want to understand my data a bit better to understand the diameter range

van_diam_summary <- summary(van_trees$diameter)

van_diam_summary

# the range is 0 -435 the mean is 11.49 and the 3rd quarter is  16.50, to visualize this, I will set my x lim to 25 to visualize the histogram again

# doing this will removes a number of rows (13438) that do no fit the criteria but gives a better visualization of diameter of 90% of the trees in the city (13438/146611 * 100)

hist_vant1 <- ggplot(van_trees, aes(x=diameter)) +
  geom_histogram(bins = 15) +
  xlim(0, 25)+
  labs(title="Tree Diameter distribution 0-25 inches") +
          ylab("frequency") +
          xlab ("diameter in")+
          theme_classic()


hist_vant1

```

## 2.1b No of tree's per Genus

Here I want to know what which species genus is the most popular in Vancouver? The relationship is between genus name and tree id.

I first need to create a new data set with two variables.Counting the number of trees for each genus and then visualize with a bar plot

```{r}

notree_spps <- van_trees %>% 
          group_by(genus_name) %>%  #group data by genus name
          tally() %>%  #count total by genus group
          arrange(desc(n))  %>%  #rearrange data to see a clearer trend
          ggplot(aes(x= reorder(genus_name,n), y = n)) +  #plot bar graph and ensure you reorder the reuslts
          geom_bar(stat = "identity")+
          coord_flip() +  #makes it easier to visualize the data 
          labs(title="Tree Genus with Most Trees") +
          ylab("number of trees") +
          xlab ("genus name") +
          theme_classic()


notree_spps

#make sure to open graph in full screen for clarity
```
Understanding the genus is key in identifying the appropriate selecting and monitoring trees. This exercise allows me to think about, is there a reason the ACER genus is mostly planted, is the tree native to Vancouver, is there more of a likely hood for survival? The data set does not have these answers but I am now curious to find out why?

## 2.1c Columns with missing na's

```{r}

van_tree_na <- van_trees %>% 
                  summarise(across(everything(), 
                   ~sum(is.na(.)))) %>%  #using acroos and everying and the summarize function to extract info needed
  pivot_longer(cols = contains("e"), names_to = "variable",     values_to = "count")%>%  #transpose  data so it easier to plot with 
  ggplot( aes(reorder(variable,count),count)) + #bar graph are a good way to visualize this data
  geom_bar(stat = "identity")+
  labs(title="Variables with missing NA values") +
  ylab("Number of NAs") +
  xlab ("Variable name") +
  coord_flip() +
  theme_bw()

van_tree_na

#to be able to select all the columns is used guidance from 
#https://www.rebeccabarter.com/blog/2020-07-09-across/
```
It is key for me to clean my data and see which key variables are missing for carrying out analysis. For instance if we don't know where the trees a planted (lat, long) or the date is was planted it becomes tricky to carry out specific analysis.

## 2.1d Current Age of tree.

This new variable will help me know how old the. the oldest tree is planted. Given the date the analysis is being carried out. 

```{r}
# created a new variable called age_tree using lubridate
age_tree <- van_trees %>% 
             mutate(age_tree =trunc(van_trees$date_planted %--%  (Sys.Date()) / years(1))) %>%  #lubridate for calculating age and trunc for rounding off see link at end of chunk
      arrange(desc(age_tree)) # arranged age from olderst to youngest tree 

age_tree

summary(age_tree$age_tree) #to assess the max min and range of age of trees

#plot a distrutbtion to visualize the data
hist_age <- ggplot(age_tree, aes(x=age_tree)) +
  geom_histogram(alpha = 0.2) +
  labs(title="Age of Tree") +
          ylab("frequency") +
          xlab ("Age in years")+
          theme_classic()

hist_age

# I used code from here to help with lubridate.
#https://datacornering.com/how-to-calculate-age-in-r/
```
I can see from the data that there a years where more tress are planted and others where less are planted.
The age variable will allow me to further explore if there are any  relationships between  age and genus planted or and  neighborhood planted, when carrying out further analysis. 

## 2.1e Which genus has tree older than 10 years and diameter greater than 10 

I did this as a bons question to test the use of filter, in this simple example I used a simple criteria to select which trees are older than 10 and have a diameter of ten


```{r}
tree_ten <- age_tree %>% 
            group_by(genus_name) %>% 
            filter(age_tree >= 10, diameter >=10)
tree_ten
```
## 2.1f Boxplot 

This again was additional as I was testing using boxplots

```{r}

#using geom_boxplox and geom_jitter to see the relationship between dimater and if the tree is planted on a curb.

boxp_1 <- ggplot(age_tree, aes(x=curb, y = diameter, color = age_tree)) +
  geom_boxplot()+
  geom_jitter()

boxp_1

```
From the data shown, trees planted by the curb have a higher diamter.

# 3 Research questions? 

1. Is there any relationship between genus, tree diameter and height? 

2. Which neighborhood has the most trees, and what genus and species are found in these neighborhoods?

3. Which year where most trees planted and where (curb,plant area,block)?

4. Are the oldest trees planted, dependent on a neighborhood,. Has the distribution of genus and spps changed over the last 30 years? 

# Comments and questions to TA's

* I was having difficulty with re-sizing labels for such a big data set, would it make sense not to use this kind of graph?

* I also attempted additional questions just so that I can learn how to carry out such tasks