---
title: "Mini Data-Analysis Deliverable 1 - Jessica Mukiri"
author: "Jessica Mukiri"
date: "19/10/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mini Data-Analysis Deliverable 2

This is my deliverable for the mini data analysis. My tasks are to:

1. Process and summarize your data (15 points)
2. Tidy your data (12.5 points)

Before starting on this task, I first need to download and load the appropriate packages

```{r }
# Packages needed
#install any of the loaded packages if not available e.g install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")

suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
```

# 1  Process and summarize of vancouver trees data

## 1.1 Research Questions

1. Is there any relationship between genus, tree diameter and height? 

2. Which neighborhood has the most trees, and what genus and species are found in these neighborhoods?

3. Which year where most trees planted and where (curb,plant area,block)?

4. Are the oldest trees planted, dependent on a neighborhood,. Has the distribution of genus and spps changed over the last 30 years? 


### 1.2 

*Question 1.Is there any relationship between genus, tree diameter and height?*

```{r}
# first I rename the Vancouver_trees data set to make it easier when running analysis 
van_trees <- vancouver_trees

#select the appropriate variables for this question

q1_sum <- van_trees %>% 
          select(genus_name,height_range_id,diameter) %>% 
          group_by(genus_name) %>% 
          summarise(mean = mean(diameter), n= n(), max = max(diameter), min = min(diameter))

```

```{r}
q1_graph <- 
```

*2.Which neighborhood has the most trees, and what genus and species are found in these neighborhoods?*

```{r}
#Compute the number of observations for at least one of your categorical variables

q2_sum <- van_trees %>% 
          group_by(genus_name,neighbourhood_name) %>% 
          summarise(n= n()) %>% 
          arrange(desc(n))
```

```{r}
q2_graph <-
```

*3.Which year where most trees planted and where (curb,plant area,block)?*

```{r}
#Based on two categorical variables, calculate two summary statistics of your choosing
q3_sum <- van_trees %>% 
          group_by(genus_name,on_street_block,curb,date_planted) %>%
          drop_na() %>% 
          separate(col = date_planted, into = c("year","month","day"),sep="-")%>% 
          mutate(year = as.numeric(year)) %>% 
          group_by(year,curb) %>% 
          summarise(n= n(), max = max(n))
```

```{r}
q3_graph <-
```

*4.Are the oldest trees planted, dependent on a neighborhood,. Has the distribution of genus and spps changed over the last 30 years?*

```{r}
#Create a categorical variable with 3 or more groups from an existing numerical variable
q4_sum <- van_trees %>% 
             mutate(age_tree =trunc(van_trees$date_planted %--%  (Sys.Date()) / years(1))) %>%  #lubridate for calculating age and trunc for rounding off see link at end of chunk
           arrange(desc(age_tree)) %>% # arranged age from oldest to youngest tree 
           mutate(age_tree = case_when( age_tree < 10 ~ "young",
                                 age_tree < 20 ~ "middle age",
                                 age_tree < 30 ~ "old",
                                  TRUE ~ "very old"))%>% 
          group_by(age_tree) %>% 
          summarise(n = n()) %>% 
          arrange(desc(n))
```

```{r}
q4_graph <- 
```


### 1.3 


# 2 Tidy data vancouver trees data

### 2.1 


### 2.2


### 2.3