---
title: "Mini Data-Analysis Deliverable 2 - Jessica Mukiri"
author: "Jessica Mukiri"
date: "29/10/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mini Data-Analysis Deliverable 3

This is my deliverable for the mini data analysis. My tasks are to:

1. Special Data Types (10 points)
2. Modelling (10 points)
3. Reading and Writing data (10 points)
4. Tidy Repository (10 points)

Before starting on this task, I first need to download and load the appropriate packages

```{r }
# Packages needed
#install any of the loaded packages if not available e.g install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")

suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(here))
```

# Research Questions

1. Which neighborhood has the most trees, and what genus and species are found in these neighborhoods?

2. Are the oldest trees planted, dependent on a neighborhood,. Has the distribution of genus and spps changed over the last 30 years?


# 1  Special Data Types

## Task: 1 Reorder factor

```{r}
## data set I am going to use
mda3_t <- vancouver_trees %>% 
                drop_na() %>% 
                separate(col = date_planted, into = c("year","month","day"),sep="-")%>% 
                mutate(year = as.numeric(year))%>%   select(tree_id,genus_name,species_name,longitude,latitude,neighbourhood_name,year)
mda3_t 

notree_spps <- mda3_t %>% 
          group_by(genus_name) %>%  #group data by genus name
          tally() %>%  #count total by genus group
          arrange(desc(n))  %>%  #rearrange data to see a clearer trend
          filter( n > 100) %>% 
          ggplot(aes(x= fct_reorder(genus_name,n), y = n)) +  #plot bar graph and ensure you reorder the results
          geom_bar(stat = "identity")+
          coord_flip() +  #makes it easier to visualize the data 
          labs(title="Tree Genus with over 100 Trees") +
          ylab("number of trees") +
          xlab ("genus name") +
          theme_classic()


notree_spps

```
This re ordring allows me to see exactly genus species is the most. The reordering  allows me to see how this is from highest to lowest. 

## Task: 3 Lubridate

```{r}
#using the lubridate data set to
van_tb <- vancouver_trees %>%
  mutate(yr = year(vancouver_trees$date_planted),
         mt = month(vancouver_trees$date_planted),
          dy = day(vancouver_trees$date_planted),
         day_year = yday(vancouver_trees$date_planted))

van_tb
```
For this task, I wanted to seperate the year month and day, I also went further to find out exactly which day of the year the trees were planted. This information can be useful to see which days and which months does tree planting occur. In future adding weather data it one could see if rainfall patterns drought influenced if more or less trees were planted (if one has daily rainfall data they can use the day column, if one as only monthly averages they can use the month column). 

# 2  Modelling

*Research Question*: Fitting the tree diameter based on date planted for Genus ACER linear regression 

*Variable of interest*: Tree diameter Y, Year X

## 2.1 Fitting a model

```{r}
van_tba <- van_tb %>% 
            filter( genus_name == "ACER")

lm_1 <-lm(diameter~ yr,van_tba)

lm_1
```

## 2.2 Model results

```{r}
lm_2 <- lm(diameter ~ I(yr-1999), data = van_tba)

lm_2 <- broom::glance(lm_2)

lm_2
```


# 3  Reading and writing data

## 3.1 Write a csv

```{r}

q1_sum <- van_tba %>% 
          select(genus_name,height_range_id,diameter) %>% 
          group_by(genus_name) %>% 
          summarise(mean = mean(diameter), n= n(), max = max(diameter), min = min(diameter), median = median(diameter), sd = sd(diameter))


write_csv(q1_sum, here::here("output","exported_file.csv"))

```


## 3.2 Write a RDS

```{r}


saveRDS(lm_1, here::here("output","model.RDS") ) 

readRDS( here::here("output","model.RDS"))
```


