---
title: "Mini Data-Analysis Deliverable 2 - Jessica Mukiri"
author: "Jessica Mukiri"
date: "29/10/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Mini Data-Analysis Deliverable 3

This is my deliverable for the mini data analysis. My tasks are to:

1. Special Data Types (10 points)
2. Modelling (10 points)
3. Reading and Writing data (10 points)
4. Tidy Repository (10 points)

Before starting on this task, I first need to download and load the appropriate packages

```{r }
# Packages needed
#install any of the loaded packages if not available e.g install.packages("devtools")
#devtools::install_github("UBC-MDS/datateachr")

suppressPackageStartupMessages(library(datateachr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lubridate))
```

# Research Questions

1. Which neighborhood has the most trees, and what genus and species are found in these neighborhoods?

2. Are the oldest trees planted, dependent on a neighborhood,. Has the distribution of genus and spps changed over the last 30 years?



# 1  Special Data Types








